Introducing ICBe: Very High Recall and Precision Event Extraction from Narratives about International Crises

### 2023-07-11 replication_paper/psrm_draft/ICBe_manuscript.qmd
r and r re-org

```{r, echo=F, cache=F, messages=F, results='hide', warnings=F, include=FALSE}
#International Crisis Behavior Events (ICBe) Dataset
knitr::opts_chunk$set(echo=F, cache=TRUE, messages=F, results='hide', warning=F, include=FALSE) #fig.width = 8, fig.height = 8, message=F, warning=F, , results=F

#Execute these lines to get bbl files for the arxiv submission
options(tinytex.clean=F)

```{r, eval=T, echo=F, cache=F}

#library(pacman)
library(googlesheets4)
library(tidyverse)
library(janitor)
library(flextable)
set_flextable_defaults(fonts_ignore=TRUE) #Warning: fonts used in `flextable` are ignored because the `pdflatex` engine is used and not `xelatex` or `lualatex`. You can avoid this warning by using the `set_flextable_defaults(fonts_ignore=TRUE)` command or use a compatible engine by defining `latex_engine: xelatex` in the YAML header of the R Markdown document.
library(ftExtra)
options(tidyverse.quiet = TRUE)
options(gargle_oauth_email = TRUE)

@@

#install.packages('tinytex')
#tinytex::install_tinytex()

```{r, echo=F}
#Notes
#https://tex.stackexchange.com/questions/83440/inputenc-error-unicode-char-u8-not-set-up-for-use-with-latex
#- \DeclareUnicodeCharacter{00A0}{ }

```{r, include=FALSE}
# options(tinytex.verbose = TRUE)
```

```{r, echo=FALSE}

```{r, echo=FALSE}

icb_long_clean <- readRDS(file=paste0(here::here(), "/replication_data/out/ICBe_V1.1_long_clean.Rds"))
codings_long_agreement <- readRDS( paste0(here::here(), "/replication_data/out/ICBe_V1.1_long_agreement.Rds") )

@@

#oh is it because break is reserved in R? 
#icb_crises <- read_csv(file=paste0(here::here(), "/data_other_datasets/icb/icb1v14.csv")) %>% janitor::clean_names()
icb_crises  <- read.csv(paste0(here::here(), "/replication_data/in/icb1v14.csv")) %>% janitor::clean_names()

#somehow break of all things is a broken name
#icb_actors <- read_csv(file=paste0(here::here(), "/data_other_datasets/icb/icb2v14.csv"))
icb_actors <- read.csv(paste0(here::here(), "/replication_data/in/icb2v14.csv"))

<!-- note: please start your introduction without including the word "Introduction" as a section heading (except for math articles in the Physical Sciences section); this heading is implied in the first paragraphs. -->

<!-- #\twocolumn -->

@@

<!-- removedcites trade @barariDemocracyTradePolicy,  @vabulasCooperationAutonomyBuilding2021 -->

<!-- removedcites @hegreIntroducingUCDPCandidate2020; ; @haltermanExtractingPoliticalEvents2020; @maozDyadicMilitarizedInterstate2019, @vabulasCooperationAutonomyBuilding2021;  -->

@@

<!-- removedcites  @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019; @brecherEscalationWarTwentieth2000; -->

This measurement paper makes the following arguments. First, there is a real-world unobserved latent concept known as international relations that can be systematically measured. Second, we propose a method for systematic large-scale measurement of the actors and behaviors in international affairs and as a proof of concept apply that method to a well-regarded and salient sample of international events known as international crises. Third, in doing so we confirm that those measurements exhibit several desirable kinds of internal and external validity and out-perform existing approaches. The article organizes that argument into XX sections. A final section concludes.

# Task Definition

We consider the measurement task of abstracting discrete events about a historical episode in international relations. The easiest way to convey the task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) alongside a mapping from each natural language sentence to discrete machine-readable abstractive events. Each event occurs over a marked time period where we observe the behaviors of a given set of actors. Our knowledge of this historical episode, including the actors involved as well as their preferences, behaviors, and and beliefs are only indirectly observed from historical records that most often take the form of unstructured natural language text. The task is to produced a structured account of the historical event by combining the unstructured corpus of historical records with informative priors about international relations. The resulting structured account of information about a historical episode can be combined with that of other events to produce a systematic account of historical political events from which we can garner novel insight given the data's high coverage, precision, and recall.

# Corpus

<!-- ^[The Online Appendix is at the [ICBEventData Github Repository](https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$).]  -->

<!-- removedcites @snyderConflictNationsBargaining1977; @gaddisExpandingDataBase1987; @georgeDeterrenceAmericanForeign1974; @brecherPatternsCrisisManagement1988; -->

```{r cuban case study precision, eval=T, results='hide', echo=F, include=T, message=F, cache=F, warning=F}

@@

ft_cuba_precision %>% save_as_image(path=here::here("replication_paper","psrm_draft","case_study_cuban_precision.png"), zoom=1) # webshot = "webshot2"
temp <- NULL

<!--- REX Remember to put in a Legend for the Icons ---->

<!-- \onecolumn -->

```{=tex}
\begin{figure}[H]
\caption{Cuban Missile Crisis (1962) - ICB Narrative vs. ICBe Events \label{fig:case_study_cuban_precision}}
\centering{\includegraphics[height=19cm]{"./case_study_cuban_precision.png"}}
\end{figure}
\clearpage
```
<!-- \twocolumn -->

# Prior Beliefs about IR, Ontological Coverage, and the Existing State of the Art

Next, we draw informative prior beliefs about the underlying process of IR that we expect to govern behavior during historical episodes and their conversion to the historical record. We organize our prior beliefs along two overarching axes, summarized in detail by Table 1.

We base our informed priors primarily on two sources of information. The first is the extensive existing measurement efforts of IR which we provide citations alongside each concept. Second, we performed preliminary natural language processing of the corpus and identified named entities and behaviors mentioned in the text. Verbs were matched to the most likely definition found in Wordnet [@millerWordNetLexicalDatabase1995], tallied, and then aggregated into a smaller number of hypernyms balancing conceptual detail and manageable sparsity for human coding (SI Appendix 1.2).

The columns compare the ontological coverage of ICBe to existing state of the art systems in production and with global coverage. They begin with our contribution ICBe, alongside other event-level datasets including CAMEO dictionary lookup-based systems (Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@daviesOrganizedViolence19892022; @sundbergIntroducingUCDPGeoreferenced2013].[^3] The final set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior2021; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018], and the Correlates of War [@sarkeesResortWar181620072010].

The rows represent the types of information we expect to find in IR and forms the basis for our proposed ontology. We employ a metaphor of a chess game, with players (polities, rebel groups, IGOs, etc.), pieces (military platforms, civilians, domains), and behaviors (think, say, do). Precise sequencing is required to capture gambits (sequences of moves) and outcomes (victory, defeat, peace, etc.), while precise geo-coding is required to understand the chessboard (medium of conflict). We find `r codings_long_agreement %>% filter(varname_normalized %>% str_detect("actor")) %>% pull(value_normalized) %>% unique() %>% length()` actors and 117 different behaviors.[^4]

[^4]: See the full codebook on Github Repository [ICBEventData](https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$).

Table 1 demonstrates that with the exception of large-scale CAMEO dictionary-based systems, our ontology improves upon the the existing state of the art quantitative datasets that ignore important information content about international interactions.[^5] In particular, we highlight two innovations in our ontology. First, we separate the chess pieces from the chess players in not only identifying what states were involved in each crisis, but distinguishing between different actors within a state. By virtue of our ontology coding military versus civilian actors and national leaders versus bureaucrats, our data can be used to explore important questions concerning topics like civilian military relations as well as evolution in what actors are engaged in crises - a topic of increasing interest as states engage in "gray zone conflict" by employing "chess pieces" like the coast guard instead of military navy. Second, we add information about the domains in which actors behave - whether in land, air, sea, space, or cyber - since they differ in their technology, tactics, geography, and purpose [@gartzkeCrossDomainDeterrenceStrategy2019]. Doing so allows researchers to identify and explain patterns in escalation conditional on the military means states use in conflict. Recent concerns about cross-domain conflict and the effect of new domains of conflict like space and cyber have made this an endeavor of increased interest to practitioners [@gannonOneIfLand2022].

[^5]: See @balaliCOfEEComprehensiveOntology2021 for a recent review of ontological depth and availability of Gold Standard example text.

```{=tex}
\clearpage
\singlespacing
```
<!-- \onecolumn -->

```{r litreview, eval=T, echo=F, results='asis', include=T, message=F, cache=F, warning=F, ft.arraystretch=0.75}

@@

# Uncomment for snapshot
# litreview_ft_psrm %>% 
#     #flextable::flextable_to_rmd()
#     save_as_image(path=here::here("replication_paper", "psrm_draft", "litreview_ft.png"), zoom=1, webshot = "webshot2") # webshot = "webshot2"

```{=tex}
\clearpage
\doublespacing
```
<!-- \twocolumn -->

```{r , eval=T, results='hide', echo=F, include=T, message=F, cache=F}

n_crisis <- ICBe_events_agreed %>% dplyr::select(crisno) %>% unique() %>% nrow()
n_sentences <- ICBe_events_agreed %>% dplyr::select(crisno, sentence_number_int_aligned) %>% unique() %>% nrow()
#n_events <- ICBe_events_agreed %>% dplyr::filter(event_number_int %in% c(1,2,3)) %>% nrow()
#n_events_cuba <- ICBe_events_agreed %>% dplyr::filter(crisno == 196 & event_number_int %in% c(1,2,3)) %>% nrow()
n_coders <- icb_long_clean %>% dplyr::select(email_id) %>% unique() %>% nrow()

@@

avg_coders_per_sentence <- icb_long_clean %>% dplyr::filter(crisno != 196) %>% dplyr::mutate(cris_sent = paste(crisno, sentence_number_int_aligned)) %>% dplyr::select(cris_sent, email_id) %>% unique() %>% dplyr::count(cris_sent) %>% dplyr::select(n) %>% lapply(mean)

# ICBe Coding Process

The ICBe ontology follows a hierarchical design philosophy where a smaller number of significant decisions are made early on and then progressively refined into more specific details [@brustIntegratingDomainKnowledge2020].[^6] Each coder was instructed to first thoroughly read the full crisis narrative and then presented with a custom graphical user interface (SI Appendix 2.1). Coders then proceeded sentence by sentence, choosing the number of events (0-3) that occurred, the highest behavior (thought, speech, or activity), a set of players, whether the means were primarily armed or unarmed, whether there was an increase or decrease in aggression (uncooperative/escalating or cooperative/de-escalating), and finally one or more non-mutually exclusive specific activities. Some additional details like location and timing information were always collected while other details were only collected if appropriate, e.g. force size, fatalities, domains, units, etc. A unique feature of the ontology is that thought, speech, and do behaviors can be nested into combinations, e.g. an offer for the U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing missiles from Turkey. Through compounding, the ontology can capture what players were said to have known, learned, or said about other specific fully described actions.

[^6]: This process quickly focuses the coder on a smaller number of relevant options while also allowing them to apply multiple tags if the sentence explicitly includes more than one or there is insufficient evidence to choose only one tag. The guided coding process also allows for the possibility that earlier coarse decisions have less error than later fine-grained decisions.

@@

```{r}
#We calculate means in the SI appendix on agreement
means_group_any <- codings_long_agreement %>% filter(value_normalized!='') %>% 
                   group_by(crisno,sentence_number_int_aligned, varname_normalized) %>% arrange(desc(percent_expert)) %>% filter(row_number()==1) %>% ungroup() %>% 
                   dplyr::select(Experts=percent_expert,Novices=percent_novice) %>% summarise_all(mean)

@@

# Internal Consistency

@@

# Case Study Selection

The remaining two qualities we seek to measure are recall and precision of ICBe events in absolute terms and relative to other existing systems. We provide full ICB narratives, ICBe coding in an easy-to-read iconographic form, and a wide range of visualizations for every case on the companion website. In this paper, we focus on two deep case studies. The first is the Cuban Missile Crisis (Figure 1) which took place primarily in the second half of 1962, involved the United States, the Soviet Union, and Cuba, and is widely known for bringing the world to the brink of nuclear war (hereafter Cuban Missiles). The second is the Crimea-Donbas Crisis (SI Appendix 4.1) which took place primarily in 2014, involved Russia, Ukraine, and NATO, and within a decade spiraled into a full-scale invasion (hereafter Crimea-Donbas). Both cases involve a superpower in crisis with a neighbor, initiated by a change from a friendly to a hostile regime, with implications for economic and military security for the superpower, risked full-scale invasion, and eventually invited intervention by opposing superpowers. We choose these cases because they are substantively significant to 20th and 21st-century international relations, widely known across scientific disciplines and popular culture, and are sufficiently brief to evaluate in depth.

# Recall

Recall measures the share of desired information recovered by a sequence of coded events, and is poorly defined for historical episodes. First, there is no genuine ground truth about what occurred, only surviving texts about it. Second, there is no \textit{a priori} guide to what information is necessary detail and what is ignorable trivia. History suffers from what is known as the Coastline Paradox [@mandelbrotFractalGeometryNature1983] --- it has a fractal dimension greater than one such that the more you zoom in, the more detail you will find about individual events and in between every two discrete events. The ICBe ontology is a proposal about what information is important, but we need an independent benchmark to evaluate whether that proposal is a good one and that allows for comparing proposals from event projects that had different goals. We need a yardstick for history.

Our strategy for dealing with both problems is a plausibly objective yardstick called a synthetic historical narrative. For both case studies, we collect a large diverse corpus of narratives spanning timelines, encyclopedia entries, journal articles, news reports, websites, and government documents. Using natural language processing (fully described in SI Appendix 3.1), we identify details that appear across multiple accounts. The more accounts that mention a detail, the more central it is to understanding the true historical episode. The theoretical motivation is that authors face word limits which force them to pick and choose which details to include, and they choose details that serve the specific context of the document they are producing. With a sufficiently large and diverse corpus of documents, we can vary the context while holding the overall episode constant and see which details tend to be invariant to context. Intuitively, a high-quality event dataset should have high recall for context invariant details both because of their broader relevance and also because they are easier to find in source material.

Synthetic historical narratives for Cuban Missiles (51 events drawn from 2020 documents) and Crimea-Donbas (30 events drawn from 971 documents) appear in Figure 2. Each row represents a detail that appeared in at least five documents along with an approximate start date, a handwritten summary, the number of documents it was mentioned in, and whether it could be identified in the text of our ICB corpus, in our ICBe events, or any of the competing systems.

From them, we draw several stylized facts. First, there is substantial variation in which details any one document will choose to include. Our ground truth ICB narratives contain 17/51 and 23/30 of the events from the synthetic narrative and include other details that are not in the synthetic narrative.[^8] Second, mentions of a detail across accounts are exponentially distributed with context-invariant details appearing dozens to hundreds of times more than context-dependent details. Third, crisis start and stop dates are arbitrary, and the historical record points to many precursor events as necessary detail for understanding later events, e.g. the U.S. was in a \textit{de facto} grey-scale war with Cuba before it invited Soviet military protection [@cormacGreyNewBlack2018] and Ukraine provided several security guarantees to Russia that were potentially undone, e.g. a long term lease on naval facilities in Crimea. Fourth, we find variation between the two cases. Cuban Missiles has a cleaner canonical end with the Soviets agreeing to withdraw missiles while Crimea-Donbas meekly ends with a second cease-fire agreement (Minsk II) but continued fighting. The canonical narrative of Cuban Missiles also includes high-level previously classified details, while the more recent Crimea-Donbas case reflects primarily public reporting.

[^8]: As the ICB narratives are intended to explain conflictual behavior in a political context, many of the missing events concern more economic components of conflict (eg. nationalizing a foreign business). Even when they occur in the context of a crisis, these events largely fall outside the sample of information on which ICBe's ontology is currently trained. Even with this limitation, ICBe is more comprehensive than the existing datasets that *do* try to code the economic dimensions of these crises. We see expanding the ontology to broader international phenomenon as a promising future implementation of our model.

```{r cuban case study recall, eval=T, results='hide', echo=F, include=T, message=F, cache=F, warning=F, ft.arraystretch=0.75}

#
#Lit review
#child=c('ICBEdataset_paper_litreviewtable.Rmd')
library(tidyverse)
library(flextable)
library(ftExtra)
flextable::set_flextable_defaults(fonts_ignore=TRUE)

```{r, eval=T, results='hide', echo=F, include=T, message=F, cache=F, warning=F, ft.arraystretch=0.75}

library(dplyr)
library(googlesheets4) #Do this high up so you get the prompt early
library(lubridate, warn.conflicts = FALSE)
case_study_crimea <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1YesAx1CkYCgrEi_WVJ9aho60HesWA6p-3XMma9IzBrI/edit?usp=sharing", 
                                sheet="final") %>% janitor::clean_names() 
dim(case_study_crimea)

```{r, recall combined, eval=T, results='hide', echo=F, include=T, message=F, cache=F, warning=F, ft.arraystretch=0.75}

ft   %>% 
    #flextable::flextable_to_rmd()
    save_as_image(path=here::here("replication_paper", "psrm_draft", "recall_cuban_and_crimea.png"), zoom=1, webshot = "webshot2") # webshot = "webshot2"

```{r, eval=T, results='hide', echo=F, include=T, message=F, cache=F, warning=F, ft.arraystretch=0.75}

#Recall Summary Statistics
#, icb_narrative_cd=icb_narrative.y, icews_cd=icews_hash
temp1_long <- temp1 %>% dplyr::select(docs=Docs, ICB_Narrative=icb_narrative, ICBe=ICBe, ICB=ICB,MIDs=MIDs,Phoenix=phoenix) %>% mutate(crisis="Cuban Missile Crisis") %>% pivot_longer(cols=-c(crisis, docs)) %>% 
              mutate(value=ifelse(!is.na(value) & !value %in% 0,1,0))

combined <- image_composite(tiger, image_scale(tiger2, "x290"), offset = "+475+650")
image_write(combined, here::here("replication_paper", "psrm_draft", "recall_cuban_and_crimea_andcounts.png"))

```{=tex}
\clearpage
\singlespacing
```
<!-- \onecolumn -->

```{=tex}
\begin{figure}[H]
\caption{Measuring Recall with Synthetic Historical Narratives \label{fig:case_study_cuban_recall}}
\centering{\includegraphics[height=18cm]{./recall_cuban_and_crimea_andcounts.png}}
\textit{Notes: Synthetic narratives combine several thousand accounts of each crisis into a single timeline of events, taking only those mentioned in at least 5 or more documents. Checkmarks represent whether that event could be hand matched to any detail in the ICB corpus, ICBe dataset, or any of the other event datasets (SI Appendix 3.2 and 3.3).}
\end{figure}
\clearpage
\doublespacing
```
<!-- \twocolumn -->

We find substantive variation in recall across systems. Recall for each increases in the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which is so noisy that it's flat to decreasing in mentions. The two episode-level datasets have low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have higher coverage, with ICEWs outperforming Terrier. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Importantly our corpus of ICB narratives has high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left little additional information on the table.

# Precision

>>>>

```{r options, echo=F, cache=F, messages=F, results='hide', warnings=F, include=FALSE}
#International Crisis Behavior Events (ICBe) Dataset
knitr::opts_chunk$set(echo=F, cache=TRUE, messages=F, results='hide', warning=F, include=FALSE) #fig.width = 8, fig.height = 8, message=F, warning=F, , results=F

#Execute these lines to get bbl files for the arxiv submission
options(tinytex.clean=F)
```

```{r package-load, eval=T, echo=F, cache=F}
#library(pacman)
library(googlesheets4)
library(tidyverse)
library(janitor)
library(flextable)
set_flextable_defaults(fonts_ignore=TRUE) #Warning: fonts used in `flextable` are ignored because the `pdflatex` engine is used and not `xelatex` or `lualatex`. You can avoid this warning by using the `set_flextable_defaults(fonts_ignore=TRUE)` command or use a compatible engine by defining `latex_engine: xelatex` in the YAML header of the R Markdown document.
library(ftExtra)
options(tidyverse.quiet = TRUE)
options(gargle_oauth_email = TRUE)

@@

#install.packages('tinytex')
#tinytex::install_tinytex()
```

```{r borked, echo=F, include=F}
#Notes
#https://tex.stackexchange.com/questions/83440/inputenc-error-unicode-char-u8-not-set-up-for-use-with-latex
#- \DeclareUnicodeCharacter{00A0}{ }
# options(tinytex.verbose = TRUE)
```

```{r load-data, echo=FALSE}
icb_long_clean <- readRDS(file=paste0(here::here(), "/replication_data/out/ICBe_V1.1_long_clean.Rds"))
codings_long_agreement <- readRDS( paste0(here::here(), "/replication_data/out/ICBe_V1.1_long_agreement.Rds") )

@@

icb_crises  <- read.csv(paste0(here::here(), "/replication_data/in/icb1v14.csv")) %>% janitor::clean_names()

icb_actors <- read.csv(paste0(here::here(), "/replication_data/in/icb2v14.csv"))
```

@@

<!-- removedcites @barariDemocracyTradePolicy; @vabulasCooperationAutonomyBuilding2021; @hegreIntroducingUCDPCandidate2020; @haltermanExtractingPoliticalEvents2020; @maozDyadicMilitarizedInterstate2019; @vabulasCooperationAutonomyBuilding2021; -->

@@

<!-- removedcites @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019; @brecherEscalationWarTwentieth2000 -->

This measurement paper proceeds in four sections that make the following arguments. First, there is a real-world unobserved latent concept known as international relations that can be systematically measured. Second, we propose a method for systematic large-scale measurement of the actors and behaviors in international affairs and as a proof of concept apply that method to a well-regarded and salient sample of international events known as international crises. Third, in doing so we confirm that those measurements exhibit several desirable kinds of internal and external validity and out-perform existing approaches. Fourth, this validation can be evaluated in detail via new event visualizations, with examples provided for case studies of the 1962 Cuban Missile Crisis and 2014 Crimea-Donbas crisis. A final section concludes.

# Identifying and measuring international relations

## Task definition

We consider the measurement task of abstracting discrete events about a historical episode in international relations. We employ a metaphor of a chess game, with players (polities, rebel groups, IGOs, etc.), pieces (military platforms, civilians, domains), and behaviors (think, say, do). Precise sequencing is required to capture gambits (sequences of moves) and outcomes (victory, defeat, peace, etc.), while precise geo-coding is required to understand the chessboard (medium of conflict). Each event occurs over a marked time period where we observe the behaviors of a given set of actors. Our knowledge of this historical episode, including the actors involved as well as their preferences, behaviors, and and beliefs are only indirectly observed from historical records that most often take the form of unstructured natural language text.

The task is to produced a structured account of the historical event by combining the unstructured corpus of historical records with informative priors about international relations. The resulting structured account of information about a historical episode can be combined with that of other events to produce a systematic account of historical political events from which we can garner novel insight given the data's high coverage, precision, and recall. The easiest way to convey the task is with an example. @fig-cuba-narrative shows a narrative account of the Cuban Missile Crisis (1962) alongside a mapping from each natural language sentence to discrete machine-readable abstractive events.

```{r cuban-case-study-precision, eval=F, results='hide', echo=F, include=T, message=F, cache=F, warning=F}
# AG temp edit: revert by setting eval=T

@@

ft_cuba_precision %>% save_as_image(path=here::here("replication_paper","psrm_draft","case_study_cuban_precision.png"), zoom=1) # webshot = "webshot2"
temp <- NULL
```

![Cuban Missile Crisis (1962) - ICB Narrative vs. ICBe Events](case_study_cuban_precision.png){#fig-cuba-narrative}

## Existing state of the art measurements

We draw informative prior beliefs about the underlying process of IR that we expect to govern behavior during historical episodes and their conversion to the historical record. We organize our prior beliefs along two overarching axes. The first is the extensive existing measurement efforts of IR, detailed below. Second, we performed preliminary natural language processing of the corpus and identified named entities and behaviors mentioned in the text. Verbs were matched to the most likely definition found in Wordnet [@millerWordNetLexicalDatabase1995], tallied, and then aggregated into a smaller number of hypernyms balancing conceptual detail and manageable sparsity for human coding (SI Appendix 1.2).

Table 1 describes the underlying structure of international crises that we hope to uncover as well as existing efforts to identify the actors and actions of international relations. The rows represent the types of information we expect to find in IR and forms the basis for our proposed ontology. The columns compare the ontological coverage of ICBe to existing state of the art systems in production and with global coverage. They begin with our contribution ICBe, alongside other event-level datasets including CAMEO dictionary lookup-based systems (Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@daviesOrganizedViolence19892022; @sundbergIntroducingUCDPGeoreferenced2013].[^3] The final set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior2021; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018], and the Correlates of War [@sarkeesResortWar181620072010].

```{=tex}
\clearpage
\singlespacing
```
```{r, eval=T, echo=F, results='asis', include=T, message=F, cache=F, warning=F, ft.arraystretch=0.75}

@@

# Uncomment for snapshot
# litreview_ft_psrm %>% 
#     #flextable::flextable_to_rmd()
#     save_as_image(path=here::here("replication_paper", "psrm_draft", "litreview_ft.png"), zoom=1, webshot = "webshot2") # webshot = "webshot2"
```

```{=tex}
\clearpage
\doublespacing
```
```{r coder-counts, eval=T, results='hide', echo=F, include=T, message=F, cache=F}
n_crisis <- ICBe_events_agreed %>% dplyr::select(crisno) %>% unique() %>% nrow()
n_sentences <- ICBe_events_agreed %>% dplyr::select(crisno, sentence_number_int_aligned) %>% unique() %>% nrow()
#n_events <- ICBe_events_agreed %>% dplyr::filter(event_number_int %in% c(1,2,3)) %>% nrow()
#n_events_cuba <- ICBe_events_agreed %>% dplyr::filter(crisno == 196 & event_number_int %in% c(1,2,3)) %>% nrow()
n_coders <- icb_long_clean %>% dplyr::select(email_id) %>% unique() %>% nrow()

@@

avg_coders_per_sentence <- icb_long_clean %>% dplyr::filter(crisno != 196) %>% dplyr::mutate(cris_sent = paste(crisno, sentence_number_int_aligned)) %>% dplyr::select(cris_sent, email_id) %>% unique() %>% dplyr::count(cris_sent) %>% dplyr::select(n) %>% lapply(mean)
```

With the exception of large-scale CAMEO dictionary-based systems, our ontology improves upon the the existing state of the art quantitative datasets that ignore important information content about international interactions.[^4] In particular, we highlight two innovations in our ontology. First, we separate the chess pieces from the chess players in not only identifying what states were involved in each crisis, but distinguishing between different actors within a state. By virtue of our ontology coding military versus civilian actors and national leaders versus bureaucrats, our data can be used to explore important questions concerning topics like civilian military relations as well as evolution in what actors are engaged in crises - a topic of increasing interest as states engage in "gray zone conflict" by employing "chess pieces" like the coast guard instead of military navy. Second, we add information about the domains in which actors behave - whether in land, air, sea, space, or cyber - since they differ in their technology, tactics, geography, and purpose [@gartzkeCrossDomainDeterrenceStrategy2019]. Doing so allows researchers to identify and explain patterns in escalation conditional on the military means states use in conflict. Recent concerns about cross-domain conflict and the effect of new domains of conflict like space and cyber have made this an endeavor of increased interest to practitioners [@gannonOneIfLand2022].

# Methodology and data

## Corpus

<!-- removedcites @snyderConflictNationsBargaining1977; @gaddisExpandingDataBase1987; @georgeDeterrenceAmericanForeign1974; @brecherPatternsCrisisManagement1988 -->

## Coding Process

The ICBe ontology follows a hierarchical design philosophy where a smaller number of significant decisions are made early on and then progressively refined into more specific details [@brustIntegratingDomainKnowledge2020].[^5] Each coder was instructed to first thoroughly read the full crisis narrative and then presented with a custom graphical user interface (SI Appendix 2.1). Coders then proceeded sentence by sentence, choosing the number of events (0-3) that occurred, the highest behavior (thought, speech, or activity), a set of players, whether the means were primarily armed or unarmed, whether there was an increase or decrease in aggression (uncooperative/escalating or cooperative/de-escalating), and finally one or more non-mutually exclusive specific activities. Some additional details like location and timing information were always collected while other details were only collected if appropriate, e.g. force size, fatalities, domains, units, etc. A unique feature of the ontology is that thought, speech, and do behaviors can be nested into combinations, e.g. an offer for the U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing missiles from Turkey. Through compounding, the ontology can capture what players were said to have known, learned, or said about other specific fully described actions. We find `r codings_long_agreement %>% filter(varname_normalized %>% str_detect("actor")) %>% pull(value_normalized) %>% unique() %>% length()` actors and 117 different behaviors.[^6]

@@

```{r means}
#We calculate means in the SI appendix on agreement
means_group_any <- codings_long_agreement %>% filter(value_normalized!='') %>% 
                   group_by(crisno,sentence_number_int_aligned, varname_normalized) %>% arrange(desc(percent_expert)) %>% filter(row_number()==1) %>% ungroup() %>% 
                   dplyr::select(Experts=percent_expert,Novices=percent_novice) %>% summarise_all(mean)

@@

# Performance comparison

## Internal consistency

@@

## Improvement over existing efforts

To evaluate our coding process relative to existing datasets, we measure the recall and precision of ICBe events in absolute terms and relative to other existing systems. Recall measures the share of desired information recovered by a sequence of coded events while precision measures the degree to which a sequence of events correctly and usefully describes the information in history. We provide full ICB narratives, ICBe coding in an easy-to-read iconographic form, and a wide range of visualizations for every case on the companion website.

Recall is poorly defined for historical episodes. First, there is no genuine ground truth about what occurred, only surviving texts about it. Second, there is no *a priori* guide to what information is necessary detail and what is ignorable trivia. History suffers from what is known as the Coastline Paradox [@mandelbrotFractalGeometryNature1983] --- it has a fractal dimension greater than one such that the more you zoom in, the more detail you will find about individual events as well as in between any two discrete events. The ICBe ontology is a proposal about what information is important, but we need an independent benchmark to evaluate whether that proposal is a good one and that allows for comparing proposals from event projects that had different goals. We need a yardstick for history.

Our strategy for dealing with both problems is a plausibly objective yardstick called a synthetic historical narrative. We collect a large diverse corpus of narratives spanning timelines, encyclopedia entries, journal articles, news reports, websites, and government documents. Using natural language processing (fully described in SI Appendix 3.1), we identify details that appear across multiple accounts. The more accounts that mention a detail, the more central it is to understanding the true historical episode. The theoretical motivation is that authors face word limits which force them to pick and choose which details to include, and they choose details that serve the specific context of the document they are producing. With a sufficiently large and diverse corpus of documents, we can vary the context while holding the overall episode constant and see which details tend to be invariant to context. Intuitively, a high-quality event dataset should have high recall for context invariant details both because of their broader relevance and also because they are easier to find in source material.

We find substantive variation in recall across existing state of the art methods. Mentions of a detail across accounts are exponentially distributed with context-invariant details appearing dozens to hundreds of times more than context-dependent details.[^8] Furthermore, crisis start and stop dates are arbitrary, and the historical record points to many precursor events as necessary detail for understanding later events. @fig-recall-cases compares ICBe's recall with that of existing datasets for the two case studies in the next section. Recall for each increases in the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which is so noisy that it's flat to decreasing in mentions. The two episode-level datasets have low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have higher coverage, with ICEWs outperforming Terrier. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Importantly our corpus of ICB narratives has high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left little additional information on the table.

![Recall comparison of two cases across existing state of the art efforts. Higher y-axis values represent higher recall and higher x-axis values represent number of times that detail is mentioned across the full corpus used to construct the synthetic narrative.](p_recall.png){#fig-recall-cases}

The second component of event measurement is precision. It does little good to recall a historical event but too vaguely (e.g. MIDs describes the Cuban Missile Crisis as a blockade, a show of force, and a stalemate) or with too much error (e.g. ICEWS records 263 "Detonate Nuclear Weapons" events between 1995-2019) to be useful for downstream applications. ICBe's ontology and coding system are designed to strike a balance so that the most important information is recovered accurately but also abstracted to a level that is still useful and interpretable. You should be able to lay out the events of a crisis on a timeline, as in Figure 3, and read off the macrostructure of an episode from each individual move. We call this visualization a crisis map, a directed graph intersected with a timeline, and provide crisis maps for every event dataset for each case study in SI Appendix 4.3 and 4.4 and for all crises on the companion website.

We further want to verify individual event codings, which we can do in the case of ICBe because each event is mapped to a specific span of text. We develop the iconography system for presenting event codings as coherent statements that can be compared side by side to the original source narrative for every case on the companion website. We further provide a stratified sample of event codings alongside their source text (SI Appendix 4.2). We find both the visualizations of macrostructure and head-to-head comparisons of ICBe codings to the raw text to strongly support the quality of ICBe.

```{r crisis-map-function, eval=T, results='hide', echo=F, include=T, message=F, cache=F, warning=F}
global_check_overlap=F
library(igraph)
library(RColorBrewer)
library(ggraph)
target_file <- paste0(here::here(),"/replication_paper/data/in/icb_manual_recoding_master_sheet.xlsx")
dictionary_actors_labels    <- readxl::read_excel(target_file, sheet="actors") %>% 
                              dplyr::select(crisno, value_normalized=value_normalized_wikidata_id, value_normalized_label) %>% dplyr::distinct() %>% na.omit()
